Biology background

### 2014-10-29 Rmarkdown/biology_background.Rmd
CpG islands and other sandy metaphors

The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). Again, there are evolutionary arguments for why these CGIs form (__TODO: sources__).

CGIs are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__). The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:

1. The moving average of GC-content is greater than $50\%$, and
2. The observed-to-expected ratio of CpGs is greater than $0.6$, and
3. The region is longer than 200 bp.

@@

The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpG canyons and CpGs in the "open sea":

- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.

__UP TO HERE__

- CpG canyons are __WHAT__ and the
- "open sea" is __WHAT AND HOW DOES THIS DIFFER FROM CpG DESERTS__.
- methylation deserts: 100 Kbp windows with the lowest 1% methylation level in the sample of interest

Another metaphor, based on methylation levels rather than CpG density, are methylation deserts. These regions are sample-dependent and were defined by \cite{Li:2012gg} as 100 Kbp windows with the lowest 1% methylation level in the given sample.

>>>>

The 20-40% of CpGs that are methylated in mammalian genomes are mostly found in clusters called _CpG islands_ (_CGIs_). CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found (__TODO: sources__). The classical definition of a CpG island, given by \cite{GardinerGarden:1987we} and used by, for example, the UCSC genome browser, is a region of the genome where:

1. The (moving) average of GC-content is greater than $50\%$, and
2. The observed-to-expected ratio of CpGs is greater than $0.6$, and
3. The region is longer than 200 bp.

@@

The "sandy/beachy" metaphors have been continued (i.e., stretched to breaking point) with various authors defining CpG island shores, CpG island shelves, CpGs in the "open sea", CpG deserts and CpG canyons.

CpG island shores, shelves and the open sea are all defined with respect to CpG islands. Specifically,

- CpG island shores: regions within 2 kb of CpG islands \cite{Irizarry:2008hg}. CpG island shores have been demonstrated to have an increased variability of CpG methylation \cite{Irizarry:2008hg}.
- CpG island shelves have been defined as regions within 2 kb of a CpG island shore \cite{Bibikova:2011fw}.
- The "open sea" contains those CpGs not classed as being in a CpG island, CpG island shore or CpG island shelf \cite{Sandoval:2011dj}.

Other metaphors, these based on methylation levels rather than CpG density, are methylation deserts \cite{Li:2012gg} and CpG canyons \cite{Jeong:2014bla}. Because these regions are defined with respect to methylation levels rather than DNA sequence, these are generally identified in a tissue-specific manner.
